---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nuggets

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/nuggets)](https://CRAN.R-project.org/package=nuggets)
[![R-CMD-check](https://github.com/beerda/nuggets/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/beerda/nuggets/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/beerda/nuggets/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/beerda/nuggets/actions/workflows/test-coverage.yaml)
[![Codecov test coverage](https://codecov.io/gh/beerda/nuggets/graph/badge.svg)](https://app.codecov.io/gh/beerda/nuggets)
[![Downloads per month](https://cranlogs.r-pkg.org/badges/nuggets)](https://cran.r-project.org/package=nuggets)
<!-- badges: end -->

Extensible R framework for
subgroup discovery ([Atzmueller (2015)](https://doi.org/10.1002/widm.1144)),
contrast patterns ([Chen (2022)](https://doi.org/10.48550/arXiv.2209.13556)),
emerging patterns ([Dong (1999)](https://doi.org/10.1145/312129.312191)) and
association rules ([Agrawal (1994)](https://www.vldb.org/conf/1994/P487.PDF)).
Both crisp (binary) and fuzzy data are supported.
It generates conditions in the form of elementary conjunctions, evaluates
them on a dataset and checks the induced sub-data for interesting statistical
properties. Currently, the package searches for implicative association rules
and conditional correlations ([Hájek (1978)](https://doi.org/10.1007/978-3-642-66943-9)).
A user-defined function may be defined to evaluate on each generated
condition to search for custom patterns.


## Installation

To install the stable version of `nuggets` from CRAN, type the following command within the R session:

``` r
install.packages("nuggets")
```

You can also install the development version of `nuggets` from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("beerda/nuggets")
```

## Examples

### Search for Implicative Rules

We start with loading of the needed packages:

```{r, message=FALSE}
library(tidyverse)
library(nuggets)
```

We are going to use the `CO2` dataset as an example:

```{r}
head(CO2)
```

First, the numeric columns need to be transformed to factors:

```{r}
d <- mutate(CO2,
            conc = cut(conc, c(-Inf, 175, 350, 675, Inf)),
            uptake = cut(uptake, c(-Inf, 17.9, 28.3, 37.12)))
head(d)
```

Then every column can be dichotomized, i.e., dummy logical columns may be created for
each factor level:

```{r}
d <- dichotomize(d)
head(d)
```

Before starting to search for the rules, it is good idea to create the vector of disjoints.
Columns with equal values in the disjoint vector will not be combined together. This will
speed-up the search as it makes no sense, e.g., to combine `Plant=Qn1` and `Plant=Qn2` in
a single condition.

```{r}
disj <- sub("=.*", "", colnames(d))
print(disj)
```

Once the data are prepared, the `dig_implications` function may be invoked. It takes the dataset
as its first parameter and a pair of "tidyselect" expressions to select the column names to appear
in the left- and right-hand side of the rule (antecedent and consequent).

```{r}
result <- dig_implications(d,
                           antecedent = !starts_with("Treatment"),
                           consequent = starts_with("Treatment"),
                           disjoint = disj,
                           min_support = 0.02,
                           min_confidence = 0.8)

result <- arrange(result, desc(support))
print(result)
```



### Custom Pattern Search

The `nuggets` package allows to execute a user-defined callback function on each generated
frequent condition. That way a custom type of patterns may be searched. The following example
replicates the search for implicative rules with the custom callback function. For that, a dataset
has to be dichotomized and the disjoint vector created as in the previous example:

```{r}
head(d)
print(disj)
```

As we want to search for implicative rules with some minimum support and confidence, we define
the variables to hold that thresholds. We also need to define a callback function that will be
called for each found frequent condition. Its purpose is to generate the rules with the
obtained condition as an antecedent:

```{r}
min_support <- 0.02
min_confidence <- 0.8

f <- function(condition, support, foci_supports) {
    conf <- foci_supports / support
    sel <- !is.na(conf) & conf >= min_confidence & !is.na(foci_supports) & foci_supports >= min_support
    conf <- conf[sel]
    supp <- foci_supports[sel]
    
    lapply(seq_along(conf), function(i) { 
      list(antecedent = format_condition(names(condition)),
           consequent = format_condition(names(conf)[[i]]),
           support = supp[[i]],
           confidence = conf[[i]])
    })
}
```

The callback function `f()` defines three arguments: `condition`, `support` and `foci_supports`.
The names of the arguments are not random. Based on the argument names of the callback function,
the searching algorithm provides information to the function. Here `condition` is a vector of indices
representing the conjunction of predicates in a condition. By the predicate we mean the column in the
source dataset. The `support` argument gets the relative frequency of the condition in the dataset.
`foci_supports` is a vector of supports of special predicates, which we call "foci" (plural of "focus"),
within the rows satisfying the condition. For implicative rules, foci are potential rule consequents.

Now we can run the digging for rules:

```{r}
result <- dig(d,
              f = f,
              condition = !starts_with("Treatment"),
              focus = starts_with("Treatment"),
              disjoint = disj,
              min_length = 1,
              min_support = min_support)
```

As we return a list of lists in the callback function, we have to flatten the first level 
of lists in the result and binding it into a data frame:

```{r}
result <- result %>%
  unlist(recursive = FALSE) %>%
  map(as_tibble) %>%
  do.call(rbind, .) %>%
  arrange(desc(support))

print(result)
```


