---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nuggets

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/nuggets)](https://CRAN.R-project.org/package=nuggets)
[![Codecov test coverage](https://codecov.io/gh/beerda/nuggets/graph/badge.svg)](https://app.codecov.io/gh/beerda/nuggets)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/nuggets)](https://cran.r-project.org/package=nuggets)
[![R-CMD-check](https://github.com/beerda/nuggets/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/beerda/nuggets/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/beerda/nuggets/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/beerda/nuggets/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

Extensible R framework for
**subgroup discovery** ([Atzmueller (2015)](https://doi.org/10.1002/widm.1144)),
**contrast patterns** ([Chen (2022)](https://doi.org/10.48550/arXiv.2209.13556)),
**emerging patterns** ([Dong (1999)](https://doi.org/10.1145/312129.312191)),
**association rules** ([Agrawal (1994)](https://www.vldb.org/conf/1994/P487.PDF)), and
**conditional correlations** ([Hájek (1978)](https://doi.org/10.1007/978-3-642-66943-9)).
Both **crisp** (Boolean, binary) and **fuzzy data** are supported.
The package generates conditions in the form of elementary conjunctions, evaluates
them on a dataset and checks the induced sub-data for interesting statistical
properties. A user-defined function may be defined to evaluate on each generated
condition to search for custom patterns.


## Installation

To install the stable version of `nuggets` from CRAN, type the following command within the R session:

``` r
install.packages("nuggets")
```

You can also install the development version of `nuggets` from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("beerda/nuggets")
```

To start using the package, load it to the R session with:

```{r}
library(nuggets)
```


## Documentation

Read the [full documentation of the nuggets package](https://beerda.github.io/nuggets/).

