---
title: "Association Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
options(tibble.width = Inf)
```



# Introduction

**Association rules** are one of the most fundamental tools in data mining.
An association rule has the form:

> *antecedent* $\Rightarrow$ *consequent*

where the *antecedent* (left-hand side) is a conjunction of predicates and
the *consequent* (right-hand side) is a single predicate. The rule expresses
that whenever the antecedent conditions are satisfied, the consequent tends
to be satisfied as well.

For example:

> `middle_age & university_edu & IT_industry` $\Rightarrow$ `high_income`

This rule states that middle-aged people with a university education working
in the IT industry tend to have a high income.

Association rules are evaluated using several quality measures:

- **Support**: the relative frequency of rows satisfying both the antecedent
  and the consequent. Higher support means the rule applies to more data.
- **Confidence**: the proportion of rows satisfying the antecedent that also
  satisfy the consequent. Higher confidence means the rule is more reliable.
- **Coverage**: the relative frequency of rows satisfying the antecedent.
- **Lift**: the ratio of the observed support to the expected support if the
  antecedent and consequent were independent. A lift greater than 1 indicates
  a positive association.

The `nuggets` package supports searching for association rules in both
**crisp** (Boolean) and **fuzzy** data through the `dig_associations()`
function.

Before using the package, the required libraries must be loaded:

```{r, message=FALSE}
library(nuggets)
library(dplyr)    # for data manipulation
```



# Data Preparation

For this tutorial, we use the built-in `CO2` dataset, which contains data
from an experiment on the cold tolerance of the grass species
*Echinochloa crus-galli*. The dataset has 84 observations and includes
information about the plant's origin (`Type`), treatment (`Treatment`),
ambient CO2 concentration (`conc`), and CO2 uptake rate (`uptake`).

```{r}
head(CO2)
```

Before searching for association rules, data must be transformed into
predicates (logical or fuzzy columns). We use the `partition()` function
for this purpose. For a detailed explanation of data preparation techniques,
see the `vignette("data-preparation")`.

## Crisp Data Preparation

We prepare a crisp version of the dataset by transforming factors into dummy
variables and numeric columns into interval-based logical predicates:

```{r}
crisp_co2 <- CO2 |>
    select(-Plant) |>
    partition(Type, Treatment) |>
    partition(conc, .method = "crisp", .breaks = c(-Inf, 200, 500, Inf)) |>
    partition(uptake, .method = "crisp", .breaks = c(-Inf, 20, 35, Inf))

head(crisp_co2, n = 3)
```

## Fuzzy Data Preparation

For a fuzzy version, we transform numeric columns into fuzzy predicates
using triangular membership functions:

```{r}
fuzzy_co2 <- CO2 |>
    select(-Plant) |>
    partition(Type, Treatment) |>
    partition(conc, .method = "triangle", .breaks = c(-Inf, 95, 500, 1000, Inf)) |>
    partition(uptake, .method = "triangle", .breaks = c(-Inf, 10, 25, 40, Inf))

head(fuzzy_co2, n = 3)
```


# Basic Association Rule Search

The simplest use of `dig_associations()` searches for all rules that
meet given minimum support and confidence thresholds.

```{r}
rules <- dig_associations(crisp_co2,
                          min_support = 0.1,
                          min_confidence = 0.8)
head(rules)
```

The result is a tibble containing found rules with their quality measures:
`antecedent`, `consequent`, `support`, `confidence`, `coverage`,
`consequent_support`, `lift`, `length`, and the contingency table
columns (`pp`, `pn`, `np`, `nn`).

For instance, the first rule represents the following association rule:

```{r, include=FALSE}
to_logical <- function(x) {
    x <- gsub("{", "", x, fixed = TRUE)
    x <- gsub("}", "", x, fixed = TRUE)
    x <- gsub(",", " & ", x, fixed = TRUE)
}
```

> `r paste0("> ", to_logical(rules$antecedent[1]), " $\\Rightarrow$ ", to_logical(rules$consequent[1]))`



# Specifying Antecedent and Consequent

In many applications, you want to constrain which predicates can appear on
each side of the rule. This is done with the `antecedent` and `consequent`
arguments, which accept
[tidyselect](https://tidyselect.r-lib.org/articles/syntax.html) expressions.

For example, to find rules that predict the `uptake` rate from all other
variables:

```{r}
rules_uptake <- dig_associations(crisp_co2,
                                 antecedent = !starts_with("uptake"),
                                 consequent = starts_with("uptake"),
                                 min_support = 0.1,
                                 min_confidence = 0.8)
head(rules_uptake, n = 3)
```


# Using the Disjoint Argument

When data is prepared with `partition()`, a single original variable is
often expanded into multiple predicates (e.g., `conc=(-Inf,200]` and
`conc=(200,500]`). These predicates from the same variable should not
appear together in the same antecedent, as their conjunction would be
contradictory or redundant. 

The `disjoint` argument prevents this. It accepts a character vector of
size equal to the number of columns in the input data such that each unique
value in the vector corresponds to a group of mutually exclusive predicates.

By default, `dig_associations()` uses `var_names()` on the column names to
create the disjoint vector automatically. This works well for data prepared
with `partition()`.

You can also provide a custom disjoint vector:

```{r}
disj <- var_names(colnames(crisp_co2))
disj

rules <- dig_associations(crisp_co2,
                          disjoint = disj,
                          min_support = 0.1,
                          min_confidence = 0.8)
head(rules, n = 3)
```


# Controlling Rule Length

The `min_length` and `max_length` arguments control the number of predicates
in the antecedent:

```{r}
# Find only rules with exactly 2 predicates in the antecedent
rules <- dig_associations(crisp_co2,
                          min_length = 2,
                          max_length = 2,
                          min_support = 0.1,
                          min_confidence = 0.8)
head(rules, n = 3)
```

Setting `min_length = 0` generates rules with an empty antecedent,
which effectively computes the support of each consequent alone.


# Limiting the Number of Results

For large datasets, the number of possible rules can be enormous. The
`max_results` argument limits the total number of rules generated:

```{r}
rules <- dig_associations(crisp_co2,
                          min_support = 0.05,
                          min_confidence = 0.6,
                          max_results = 5)
nrow(rules)
```


# Fuzzy Association Rules

When the data contains fuzzy predicates (numeric columns with values in
[0, 1]), `dig_associations()` computes fuzzy support using a
**t-norm** for conjunction. The `t_norm` argument specifies which t-norm
to use:

- `"goguen"` (default): the product t-norm - multiplies membership degrees
- `"goedel"`: the minimum t-norm - takes the minimum of membership degrees
- `"lukas"`: the Łukasiewicz t-norm = max(0, a + b - 1)

```{r}
# Fuzzy rules using the product t-norm (default)
fuzzy_rules <- dig_associations(fuzzy_co2,
                                antecedent = !starts_with("uptake"),
                                consequent = starts_with("uptake"),
                                min_support = 0.05,
                                min_confidence = 0.6,
                                t_norm = "goguen")
head(fuzzy_rules, n = 3)
```

The choice of t-norm affects how strictly the conjunction is evaluated.
The Gödel t-norm is the least strict (produces higher support values),
while the Łukasiewicz t-norm is the most strict. Note also that handling fuzzy data is
generally much more slower and memory demanding than crisp data, especially 
for large datasets.


# Computing Additional Interest Measures

The `add_interest()` function computes additional interestingness measures
for association rules beyond the basic support, confidence, and lift.
It uses the contingency table columns (`pp`, `pn`, `np`, `nn`) that are
automatically included in the output of `dig_associations()`.

```{r}
# Add selected interest measures
rules_enriched <- rules_uptake |>
    add_interest(measures = c("conviction", "leverage", "jaccard"))
rules_enriched |>
    select(antecedent, consequent, confidence, conviction, leverage, jaccard) |>
    head(n = 3)
```

You can compute all available measures at once by omitting the `measures`
argument:

```{r}
rules_all_measures <- rules_uptake |>
    add_interest()
colnames(rules_all_measures)
```

See the documentation of `add_interest()` for a complete list of supported
quality measures:

```{r, eval=FALSE}
?add_interest
```



## Smoothing

Some interest measures may be undefined when contingency table counts are
zero (e.g., division by zero). The `smooth_counts` argument applies Laplace
smoothing to the counts before computing the measures:

```{r}
rules_smoothed <- rules_uptake |>
    add_interest(measures = c("odds_ratio", "conviction"),
                 smooth_counts = 0.5)
rules_smoothed |>
    select(antecedent, consequent, confidence, odds_ratio, conviction) |>
    head(n = 3)
```

## GUHA Quantifiers

`add_interest()` also supports GUHA (General Unary Hypothesis Automaton)
quantifiers, including statistical tests based on the binomial distribution:

```{r}
rules_guha <- rules_uptake |>
    add_interest(measures = c("dfi", "fe", "lci"),
                 p = 0.5)
rules_guha |>
    select(antecedent, consequent, confidence, dfi, fe, lci) |>
    head(n = 3)
```

The `p` parameter represents the null-hypothesis probability used in the
binomial-test-based quantifiers (`lci`, `uci`, `dlci`, `duci`, `lce`, `uce`).


# Excluding Tautologies from the Search

In real-world datasets, some rules are trivially true or nearly so — for
example, rules that follow directly from the structure of the data (e.g.,
`engine_type=electric => fuel_type=electricity`). Such rules are called
**tautologies**. While technically valid, they are often uninteresting and
can clutter results or slow down the search.

The `excluded` argument of `dig_associations()` allows you to specify a
list of predicate combinations that should never appear together in
generated rules. Note that the term "tautology" here is used broadly - any rule
you consider uninteresting (regardless of whether its confidence is exactly 1)
can be excluded this way.


## Finding Tautologies with `dig_tautologies()`

The `dig_tautologies()` function is specifically designed to find rules
with very high confidence. It searches iteratively, using tautologies found
in earlier iterations to prune the search space in later iterations. For a
detailed description, see the `vignette("data-preparation")`.

```{r}
tautologies <- dig_tautologies(crisp_co2,
                               antecedent = everything(),
                               consequent = everything(),
                               min_confidence = 0.95,
                               min_support = 0.05,
                               max_length = 2)
tautologies
```


## Using Tautologies to Filter Association Rules

Once tautologies are identified, convert them to the format expected by the
`excluded` argument using `parse_condition()`, and pass them to
`dig_associations()`:

```{r}
# Convert tautologies to the excluded format
excluded_conds <- parse_condition(tautologies$antecedent,
                                  tautologies$consequent)

# Search for rules while excluding tautological patterns
rules_filtered <- dig_associations(crisp_co2,
                                   antecedent = !starts_with("uptake"),
                                   consequent = starts_with("uptake"),
                                   excluded = excluded_conds,
                                   min_support = 0.1,
                                   min_confidence = 0.8)
rules_filtered
```

By excluding known tautologies, the search focuses on genuinely interesting
patterns and can run significantly faster on large datasets.


## Manually Specifying Exclusions

You can also construct the `excluded` list manually. Each element is a
character vector of elements that should not appear together in a rule:

```{r}
# Filter out rules by tautology "Treatment=chilled => Type=Mississippi"
manual_excluded <- list(c("Treatment=chilled", "Type=Mississippi"))

rules_manual <- dig_associations(crisp_co2,
                                 antecedent = !starts_with("uptake"),
                                 consequent = starts_with("uptake"),
                                 excluded = manual_excluded,
                                 min_support = 0.1,
                                 min_confidence = 0.8)
rules_manual
```

This is useful when you have domain knowledge about which combinations are
trivial or undesirable, without needing to run `dig_tautologies()` first.


# Summary

This vignette demonstrated how to search for association rules using the
`nuggets` package:

1. **Data preparation** with `partition()` transforms raw data into crisp
   or fuzzy predicates suitable for rule mining (see `vignette("data-preparation")`
   for details).

2. **Basic search** with `dig_associations()` finds rules meeting minimum
   support and confidence thresholds.

3. **Controlling the search space**: use `antecedent`/`consequent` to
   constrain which predicates appear on each side, `disjoint` to prevent
   contradictory combinations via `var_names()`, and `min_length`/`max_length`
   to control rule complexity.

4. **Fuzzy rules** extend the approach to graded membership using t-norms
   (`"goguen"`, `"goedel"`, `"lukas"`).

5. **Interest measures** can be added with `add_interest()` to evaluate
   rules from multiple perspectives (conviction, leverage, Jaccard, GUHA
   quantifiers, and many more).

6. **Excluding tautologies** with the `excluded` argument and
   `dig_tautologies()` removes trivial or uninteresting rules and speeds up
   the search.

For further details, consult the function documentation:
`dig_associations()`, `add_interest()`, `dig_tautologies()`,
`parse_condition()`, `partition()`, `var_names()`.
