---
title: "Association Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(nuggets)
library(dplyr)

options(tibble.width = Inf)
```


# Introduction

**Association rules** are one of the most fundamental tools in data mining.
An association rule has the form:

> *antecedent* $\Rightarrow$ *consequent*

where the *antecedent* (left-hand side) is a conjunction of predicates and
the *consequent* (right-hand side) is a single predicate. The rule expresses
that whenever the antecedent conditions are satisfied, the consequent tends
to be satisfied as well.

For example:

> `middle_age & university_edu & IT_industry` $\Rightarrow$ `high_income`

This rule states that middle-aged people with a university education working
in the IT industry tend to have a high income.

Association rules are evaluated using several quality measures:

- **Support**: the relative frequency of rows satisfying both the antecedent
  and the consequent. Higher support means the rule applies to more data.
- **Confidence**: the proportion of rows satisfying the antecedent that also
  satisfy the consequent. Higher confidence means the rule is more reliable.
- **Coverage**: the relative frequency of rows satisfying the antecedent.
- **Lift**: the ratio of the observed support to the expected support if the
  antecedent and consequent were independent. A lift greater than 1 indicates
  a positive association.

The `nuggets` package supports searching for association rules in both
**crisp** (Boolean) and **fuzzy** data through the `dig_associations()`
function.


# Data Preparation

For this tutorial, we use the built-in `CO2` dataset, which contains data
from an experiment on the cold tolerance of the grass species
*Echinochloa crus-galli*. The dataset has 84 observations and includes
information about the plant's origin (`Type`), treatment (`Treatment`),
ambient CO2 concentration (`conc`), and CO2 uptake rate (`uptake`).

```{r}
head(CO2)
```

Before searching for association rules, data must be transformed into
predicates (logical or fuzzy columns). We use the `partition()` function
for this purpose. For a detailed explanation of data preparation techniques,
see the `vignette("data-preparation")`.

## Crisp Data Preparation

We prepare a crisp version of the dataset by transforming factors into dummy
variables and numeric columns into interval-based logical predicates:

```{r}
crisp_co2 <- CO2 |>
    select(-Plant) |>
    partition(Type, Treatment) |>
    partition(conc, .method = "crisp", .breaks = c(-Inf, 200, 500, Inf)) |>
    partition(uptake, .method = "crisp", .breaks = c(-Inf, 20, 35, Inf))

head(crisp_co2, n = 3)
```

## Fuzzy Data Preparation

For a fuzzy version, we transform numeric columns into fuzzy predicates
using triangular membership functions:

```{r}
fuzzy_co2 <- CO2 |>
    select(-Plant) |>
    partition(Type, Treatment) |>
    partition(conc, .method = "triangle", .breaks = c(-Inf, 95, 500, 1000, Inf)) |>
    partition(uptake, .method = "triangle", .breaks = c(-Inf, 10, 25, 40, Inf))

head(fuzzy_co2, n = 3)
```


# Basic Association Rule Search

The simplest use of `dig_associations()` searches for all rules that
meet given minimum support and confidence thresholds.

```{r}
rules <- dig_associations(crisp_co2,
                          min_support = 0.1,
                          min_confidence = 0.8)
rules
```

The result is a tibble containing found rules with their quality measures:
`antecedent`, `consequent`, `support`, `confidence`, `coverage`,
`consequent_support`, `lift`, `length`, and the contingency table
columns (`pp`, `pn`, `np`, `nn`).


# Specifying Antecedent and Consequent

In many applications, you want to constrain which predicates can appear on
each side of the rule. This is done with the `antecedent` and `consequent`
arguments, which accept
[tidyselect](https://tidyselect.r-lib.org/articles/syntax.html) expressions.

For example, to find rules that predict the `uptake` rate from all other
variables:

```{r}
rules_uptake <- dig_associations(crisp_co2,
                                 antecedent = !starts_with("uptake"),
                                 consequent = starts_with("uptake"),
                                 min_support = 0.1,
                                 min_confidence = 0.8)
rules_uptake
```


# Using the Disjoint Argument

When data is prepared with `partition()`, a single original variable is
often expanded into multiple predicates (e.g., `conc=(-Inf,200]` and
`conc=(200,500]`). These predicates from the same variable should not
appear together in the same antecedent, as their conjunction would be
contradictory or redundant.
The `disjoint` argument prevents this. By default, `dig_associations()`
uses `var_names()` on the column names to create groups of mutually
exclusive predicates.

You can also provide a custom disjoint vector:

```{r}
disj <- var_names(colnames(crisp_co2))
head(disj)

rules_disj <- dig_associations(crisp_co2,
                               antecedent = !starts_with("uptake"),
                               consequent = starts_with("uptake"),
                               disjoint = disj,
                               min_support = 0.1,
                               min_confidence = 0.8)
rules_disj
```


# Controlling Rule Length

The `min_length` and `max_length` arguments control the number of predicates
in the antecedent:

```{r}
# Find only rules with exactly 2 predicates in the antecedent
rules_len2 <- dig_associations(crisp_co2,
                               antecedent = !starts_with("uptake"),
                               consequent = starts_with("uptake"),
                               min_length = 2,
                               max_length = 2,
                               min_support = 0.1,
                               min_confidence = 0.8)
rules_len2
```

Setting `min_length = 0` generates rules with an empty antecedent,
which effectively computes the support of each consequent alone.


# Limiting the Number of Results

For large datasets, the number of possible rules can be enormous. The
`max_results` argument limits the total number of rules generated:

```{r}
rules_limited <- dig_associations(crisp_co2,
                                  antecedent = !starts_with("uptake"),
                                  consequent = starts_with("uptake"),
                                  min_support = 0.05,
                                  min_confidence = 0.6,
                                  max_results = 20)
rules_limited
```


# Fuzzy Association Rules

When the data contains fuzzy predicates (numeric columns with values in
$[0, 1]$), `dig_associations()` computes fuzzy support using a
**t-norm** for conjunction. The `t_norm` argument specifies which t-norm
to use:

- `"goguen"` (default): the product t-norm — multiplies membership degrees
- `"goedel"`: the minimum t-norm — takes the minimum of membership degrees
- `"lukas"`: the Łukasiewicz t-norm — $\max(0, a + b - 1)$

```{r}
# Fuzzy rules using the product t-norm (default)
fuzzy_rules <- dig_associations(fuzzy_co2,
                                antecedent = !starts_with("uptake"),
                                consequent = starts_with("uptake"),
                                min_support = 0.05,
                                min_confidence = 0.6,
                                t_norm = "goguen")
fuzzy_rules
```

Comparing different t-norms:

```{r}
# Using the Goedel (minimum) t-norm
fuzzy_rules_goedel <- dig_associations(fuzzy_co2,
                                       antecedent = !starts_with("uptake"),
                                       consequent = starts_with("uptake"),
                                       min_support = 0.05,
                                       min_confidence = 0.6,
                                       t_norm = "goedel")
nrow(fuzzy_rules_goedel)
```

The choice of t-norm affects how strictly the conjunction is evaluated.
The Gödel t-norm is the least strict (produces higher support values),
while the Łukasiewicz t-norm is the most strict.


# Computing Additional Interest Measures

The `add_interest()` function computes additional interestingness measures
for association rules beyond the basic support, confidence, and lift.
It uses the contingency table columns (`pp`, `pn`, `np`, `nn`) that are
automatically included in the output of `dig_associations()`.

```{r}
# Add selected interest measures
rules_enriched <- rules_uptake |>
    add_interest(measures = c("conviction", "leverage", "jaccard"))
rules_enriched |>
    select(antecedent, consequent, confidence, conviction, leverage, jaccard)
```

You can compute all available measures at once by omitting the `measures`
argument:

```{r}
rules_all_measures <- rules_uptake |>
    add_interest()
colnames(rules_all_measures)
```

## Smoothing

Some interest measures may be undefined when contingency table counts are
zero (e.g., division by zero). The `smooth_counts` argument applies Laplace
smoothing to the counts before computing the measures:

```{r}
rules_smoothed <- rules_uptake |>
    add_interest(measures = c("odds_ratio", "conviction"),
                 smooth_counts = 0.5)
rules_smoothed |>
    select(antecedent, consequent, confidence, odds_ratio, conviction)
```

## GUHA Quantifiers

`add_interest()` also supports GUHA (General Unary Hypothesis Automaton)
quantifiers, including statistical tests based on the binomial distribution:

```{r}
rules_guha <- rules_uptake |>
    add_interest(measures = c("dfi", "fe", "lci"),
                 p = 0.5)
rules_guha |>
    select(antecedent, consequent, confidence, dfi, fe, lci)
```

The `p` parameter represents the null-hypothesis probability used in the
binomial-test-based quantifiers (`lci`, `uci`, `dlci`, `duci`, `lce`, `uce`).


# Parallel Computation

For large datasets, you can speed up the search using multiple threads
with the `threads` argument:
 
```{r, eval=FALSE}
rules_parallel <- dig_associations(crisp_co2,
                                   antecedent = !starts_with("uptake"),
                                   consequent = starts_with("uptake"),
                                   min_support = 0.05,
                                   min_confidence = 0.6,
                                   threads = 4)
```


# Summary

This vignette demonstrated how to search for association rules using the
`nuggets` package:
 
1. **Data preparation** with `partition()` transforms raw data into crisp
   or fuzzy predicates suitable for rule mining (see `vignette("data-preparation")`
   for details).

2. **Basic search** with `dig_associations()` finds rules meeting minimum
   support and confidence thresholds.

3. **Controlling the search space**: use `antecedent`/`consequent` to
   constrain which predicates appear on each side, `disjoint` to prevent
   contradictory combinations via `var_names()`, and `min_length`/`max_length`
   to control rule complexity.

4. **Fuzzy rules** extend the approach to graded membership using t-norms
   (`"goguen"`, `"goedel"`, `"lukas"`).

5. **Interest measures** can be added with `add_interest()` to evaluate
   rules from multiple perspectives (conviction, leverage, Jaccard, GUHA
   quantifiers, and many more).

6. **Performance** can be managed with `max_results` and `threads`.

For further details, consult the function documentation:
`dig_associations()`, `add_interest()`, `partition()`, `var_names()`.
