---
title: "Comparison of nuggets and arules performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of nuggets and arules performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    echo = FALSE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4
)
options(tibble.width = Inf)

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(kableExtra)

preprocess <- function(d) {
    d |>
        pivot_longer(cols = c("nuggets", "arules_apriori", "arules_eclat"),
                     names_to = "method",
                     values_to = "time")  |>
        mutate(time = time / 1e6,  # convert nanoseconds to milli-seconds
               method = dplyr::recode(method,
                                "nuggets" = "nuggets",
                                "arules_apriori" = "apriori (arules)",
                                "arules_eclat" = "eclat (arules)"),
               method = factor(method)) 
}


two_fig <- function(d, x, xname, title) {
    nice_colnames <- c(cols = "# of data columns",
                        rows = "# of data rows",
                        min_support = "min supp",
                        min_confidence = "min conf",
                        max_length = "max ante length")
    fixed <- list()
    for (col in c("rows", "cols", "min_support", "min_confidence", "max_length")) {
        if (length(unique(d[[col]])) == 1) {
            fixed[[col]] <- unique(d[[col]])
        }
    }
    subtitle <- paste0("(",
                       paste(nice_colnames[names(fixed)], fixed, sep = " = ", collapse = ", "),
                       ")")
    
    p1 <- ggplot(d) +
        aes(x = !!x, y = time, color = method) +
        geom_point() +
        geom_line() +
        labs(title = "linear scales",
             x = xname,
             y = "Time [ms]",
             color = "Method")
    
    p2 <- ggplot(d) +
        aes(x = !!x, y = time, color = method) +
        geom_point() +
        geom_line() +
        scale_x_log10() +
        scale_y_log10() +
        labs(title = "log scales",
             x = xname,
             y = "Time [ms]",
             color = "Method")
    
    p1 + p2 +
        plot_annotation(title = title, subtitle = subtitle) +
        plot_layout(ncol = 2, guides = "collect") & theme(legend.position = "bottom") 
    
}


tab <- function(d, title) {
    d |>
        select(rows, cols, method, time) |>
        mutate(time = round(time, 2)) |>
        pivot_wider(names_from = method, values_from = time) |>
        relocate(`eclat (arules)`, .after = `nuggets`) |>
        arrange(rows, cols) |> 
        #kable(caption = title) |> 
        kable() |> 
        kable_styling(full_width = TRUE) |> 
        add_header_above(c(" " = 1, " " = 1, "Time [ms]" = 3))
}

data <- readRDS("comparison-with-arules.rds")

dense_rows <- data$dense_rows |> preprocess()
dense_cols <- data$dense_cols |> preprocess()
sparse_rows <- data$sparse_rows |> preprocess()
sparse_cols <- data$sparse_cols |> preprocess()
```


# Introduction

This vignette compares the performance of the `nuggets` and `arules` packages 
in R on the task of *association rule mining*. The comparison focuses on the
time needed to discover association rules *from Boolean datasets* of varying
sizes and sparsity levels. The goal is to provide a rough comparison of 
computational power rather than a full comparison of package functionality,
so advanced features, filtering options, and other factors that may affect
practical performance are not considered here.

For reproducibility, the
[benchmark script used in this vignette](https://github.com/beerda/nuggets/tree/main/misc/performance/nuggets_vs_arules)
is available in the package repository on GitHub.

# Materials and Methods

A series of experiments were conducted to evaluate the performance of the
`nuggets` and `arules` packages. For `arules`, two different algorithms were
evaluated: the Apriori algorithm (`apriori()`) and and the Eclat algorithm 
(`eclat()`). For `nuggets`, the `dig_associations()` function was used to 
discover association rules. The experiments were designed to measure the 
execution time of each method under different conditions, including varying the
number of rows and columns in the datasets, as well as the sparsity of the data.

The test datasets were randomly generated with binary values (TRUE/FALSE)
and varying numbers of rows and columns. The sparsity of the data was controlled
by adjusting the probability of TRUE values in the dataset.

Specifically, the following parameters were varied in the experiments:

- number of rows: 10^3^, 10^4^, 10^5^, 10^6^
- number of columns: 10, 20, 30, 50, 80
- the probability of TRUE values: 0.5 (dense datasets) and 0.1 (sparse datasets)

The other parameters were kept constant across all experiments:

- the minimum support threshold: 0.001
- the minimum confidence threshold: 0.5 (dense datasets) and 0.1 (sparse datasets)
- the maximum length of antecedents: 3

Each experiment was repeated 5 times to ensure the reliability of the 
results, and the average execution time was recorded.
The results are visualized using both linear and logarithmic scales to provide
insights into the performance characteristics of each method. All experiments
were conducted on  `r data$cpu` (`r data$cache` cache) with 
`r round(data$ram, 1)` GB of RAM available.


# Results

## Dense data: varying number of rows

```{r}
tab(dense_rows, "Execution time on dense data with varying number of rows")
```

```{r}
two_fig(dense_rows, 
        sym("rows"),
        "Number of data rows",
        "Execution time on dense data with varying number of rows")
```

## Dense data: varying number of columns

```{r}
tab(dense_cols, "Execution time on dense data with varying number of columns")
```

```{r}
two_fig(dense_cols, 
        sym("cols"),
        "Number of data columns",
        "Execution time on dense data with varying number of columns")
```

## Sparse data: varying number of rows

```{r}
tab(sparse_rows, "Execution time on sparse data with varying number of rows")
```

```{r}
two_fig(sparse_rows, 
        sym("rows"),
        "Number of data rows",
        "Execution time on sparse data with varying number of rows")
```

## Sparse data: varying number of columns

```{r}
tab(sparse_cols, "Execution time on sparse data with varying number of columns")
```

```{r}
two_fig(sparse_cols, 
        sym("cols"),
        "Number of data columns",
        "Execution time on sparse data with varying number of columns")
```


# Discussion

The results show that the relative performance of `nuggets` and `arules` depends
strongly on the structure of the data, particularly on its dimensionality and
sparsity. In general, `nuggets` performs best when the number of predicates is
small to moderate, whereas `arules` becomes increasingly competitive as the data
become sparser and the number of predicates grows.

## Dense data

For dense datasets with a fixed number of rows, all methods become slower as
the number of columns increases. Nevertheless, nuggets remains the fastest 
across the entire range considered. `arule`'s `eclat()` is consistently slower
than  `nuggets` but faster than `arule`'s `apriori()`. Although the 
computational cost rises rapidly for all three methods as the number of
predicates grows, the relative ordering does not change in the dense-data
experiments.

## Sparse data

For sparse datasets with a fixed number of columns, the same general ranking
is observed as in the dense case. Execution time increases with the number of
rows, but `nuggets` remains the fastest method throughout, with `arule`'s 
`eclat()` second and `apriori()` third. The results suggest that `nuggets` also
scales favorably with increasing dataset size in sparse data, at least when
the number of columns remains small.

A different pattern emerges when the number of columns increases in sparse
datasets. For smaller and moderate numbers of columns, `nuggets` remains the
fastest method. However, the gap between methods narrows as dimensionality
grows. At higher numbers of columns, `arule`'s `apriori()` overtakes both 
Eclat-based approaches (yes, `nuggets` is inspirred with the Eclat algorithm 
internally) and becomes the fastest method, while `nuggets` becomes the slowest
among the three. This indicates that sparse, high-dimensional data favor the
search strategy used by `apriori()` over the recursion-based approaches 
used by `eclat()` and `nuggets`.

Overall, the experiments indicate that `nuggets` performs best in low- to
medium-dimensional settings, both dense and sparse, and shows very good
scalability with increasing numbers of rows. In contrast, for sparse datasets
with many predicates, `apriori()` becomes more efficient and ultimately provides
the best performance.

## Why `nuggets` performs well

A likely explanation for the strong performance of `nuggets` is its highly 
optimized implementation of conjunction computation and support counting.
These operations are central to rule discovery, and in `nuggets` they are
accelerated using SIMD instructions. This makes the evaluation of candidate
conjunctions particularly fast and is reflected in the consistently strong
performance of `nuggets` in experiments with increasing numbers of rows. In such
settings, the amount of data to be processed grows, but the low-level
optimizations allow `nuggets` to maintain a substantial runtime advantage. All
nuggets optimizations are available with the default compiler directives 
recommended by CRAN, without requiring any non-standard package installation 
settings.


# Summary

The results show that `nuggets` is particularly effective for low- to
medium-dimensional data, where its optimized implementation provides 
consistently strong performance. For sparse data with many predicates, however,
`arules`, especially `apriori()`, becomes more advantageous. Overall, the
comparison highlights that the best choice depends mainly on data sparsity and
dimensionality.

For additional information on the `nuggets` package, see:

- `vignette("nuggets")` for an overview of the package and its main workflows,
- `vignette("association-rules")` for a specialized pattern family based on the
  `dig_associations()` function,
- `vignette("conditional-correlations")` for subgroup-based correlation
  analysis on numeric variables,
- `vignette("contrast-patterns")` for subgroup-based statistical comparisons
  of numeric variables,
- `vignette("custom-patterns")` for defining custom pattern types.

