---
title: "Custom Pattern Search with dig()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Pattern Search with dig()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r, include = FALSE}
options(tibble.width = Inf)
```



# Introduction

`dig()` is the general function behind custom pattern search in the `nuggets`
package. It searches for patterns of a custom type by generating conditions as
elementary conjunctions of predicates and by executing a user-defined callback
function on each generated condition.

This makes `dig()` the low-level building block behind more specialized
functions such as `dig_associations()` and `dig_correlations()`. Use it when
you want to keep the search over conditions, but define your own evaluation
logic.

This vignette focuses on how to use the `dig()` function: how to write the
callback function and how to control the search. For preparation of crisp and
fuzzy predicates, see `vignette("data-preparation")`.

Examples in this vignette require loading the following packages:

```{r, message = FALSE}
library(nuggets)
library(dplyr)     # for data manipulation
```



# A Small Working Dataset

`dig()` expects a matrix or data frame whose columns are logical predicates or
numeric fuzzy predicates. We will use `iris` and prepare a small predicate data
set that is rich enough for the examples below:

```{r}
dig_iris <- iris |>
    partition(Species) |>
    partition(Sepal.Length:Petal.Width, .method = "crisp", .breaks = 3)

disj <- var_names(colnames(dig_iris))
head(dig_iris, n = 3)
```

The preparation step is intentionally brief here; the dedicated
`vignette("data-preparation")` explains `partition()`, fuzzy predicates,
breakpoints, and `var_names()` in detail.



# Terminology: Condition and Focus

`dig()` works with two kinds of predicates:

- **Condition**: an elementary conjunction generated from columns selected by
  `condition`.
- **Focus**: predicates selected by `focus` and tested within each generated
  condition.

You can think of the scheme as:

> *condition* `|` *focus set*

For example, if a generated condition is
`Sepal.Length=(-Inf,5.5] & Petal.Width=(-Inf,0.8]` and the focus set contains
the species predicates, then `dig()` can tell your callback how often each
species occurs inside that condition.

When `focus` is used, `dig()` can pass contingency-table entries to the
callback:

- `pp`: rows satisfying both the condition and a focus,
- `pn`: rows satisfying the condition but not the focus,
- `np`: rows satisfying the focus but not the condition,
- `nn`: rows satisfying neither.

If you do not need foci, leave `focus = NULL` and write a callback that only
uses condition-level information.



# A Simple `dig()` Call

The simplest pattern search uses a callback that receives a generated condition
and computes its own output. The callback only needs to declare the arguments
it actually uses.

In the next example, the callback returns the formatted condition, its support,
and the support of all species predicates inside that condition:

```{r}
simple_callback <- function(condition, support, pp) {
    c(
        list(
            condition = format_condition(names(condition)),
            condition_support = support
        ),
        as.list(pp / nrow(dig_iris))
    )
}

simple_result <- dig(
    x = dig_iris,
    f = simple_callback,
    condition = !starts_with("Species"),
    focus = starts_with("Species"),
    disjoint = disj,
    min_length = 1,
    max_length = 1,
    min_support = 0.2,
    min_focus_support = 0
) |>
    bind_rows()

simple_result
```

This illustrates the basic workflow:

1. choose which predicates may form conditions,
2. optionally choose focus predicates,
3. define a callback,
4. let `dig()` enumerate conditions and collect callback results.



# What the Callback Function Can Receive

The callback function `f` may declare any subset of the following arguments.
`dig()` detects which arguments are present and computes only those values.

- `condition`: named integer vector of column indices representing the generated
  condition.
- `sum`: number of rows satisfying the condition for logical data, or the sum
  of truth degrees for fuzzy data.
- `support`: relative frequency of the condition.
- `indices`: row indices satisfying the condition in crisp searches.
- `weights`: per-row truth degrees of the condition in fuzzy searches.
- `pp`, `pn`, `np`, `nn`: contingency-table entries for the remaining foci.
- `foci_supports`: deprecated focus-support argument kept for backward
  compatibility.

In practice:

- use `condition` when you need predicate names,
- use `support` or `sum` for condition-level filtering or ranking,
- use `indices` when you want to compute something on the original rows,
- use `weights` for custom fuzzy summaries,
- use `pp`, `pn`, `np`, and `nn` when your pattern depends on foci.



# Main Arguments of `dig()`

The `dig()` function exposes the full condition-generation engine. The most
important arguments are:

- `x`: matrix or data frame containing logical or fuzzy predicates.
- `f`: callback function executed for each generated condition.
- `condition`: tidyselect expression specifying which columns of `x` may appear
  in generated conditions.
- `focus`: tidyselect expression specifying which columns of `x` are evaluated
  as foci within each condition.
- `disjoint`: vector defining groups of mutually exclusive predicates; predicates
  from the same group are not combined in one condition.
- `excluded`: list of known implications (axioms) used for pruning; each element
  is a character vector where all but the last value form the antecedent and
  the last value is the consequent.
- `min_length` and `max_length`: minimum and maximum number of predicates in a
  generated condition.
- `min_support` and `max_support`: support range for conditions that should
  trigger the callback.
- `min_focus_support`: minimum support of a focus inside a condition.
- `min_conditional_focus_support`: minimum conditional support of a focus within
  a condition.
- `filter_empty_foci`: if `TRUE`, skip callback calls for conditions that have
  no remaining foci after focus filtering.
- `t_norm`: conjunction operator for fuzzy data (`"goedel"`, `"goguen"`, or
  `"lukas"`).
- `max_results`: maximum number of callback results to store before stopping.
- `verbose`: print progress messages.
- `threads`: number of threads used for the search.
- `error_context`: helper for wrapper authors so that errors refer to caller-side
  argument names; most interactive uses can ignore it.

Two arguments are especially important for performance: `min_support` and
`max_length`. Together they strongly control the size of the search space.



# A More Advanced Example: Fixed-Variable Correlations

`dig_correlations()` searches over both generated conditions and combinations of
numeric variables. A simpler custom variant can be built directly with `dig()`
when the two variables are fixed in advance and only the condition should vary.

Here we search for conditions under which `Sepal.Length` and `Petal.Length`
correlate strongly. The callback receives `indices`, uses them to select the
corresponding rows from the original `iris` data, and runs `cor.test()` on that
sub-data:

```{r}
fixed_xy <- iris[, c("Sepal.Length", "Petal.Length")]

correlation_callback <- function(condition, support, indices) {
    if (length(indices) < 10) {
        return(NULL)
    }

    fit <- cor.test(
        fixed_xy$Sepal.Length[indices],
        fixed_xy$Petal.Length[indices],
        method = "pearson"
    )

    list(
        condition = format_condition(names(condition)),
        support = support,
        correlation = unname(fit$estimate),
        p_value = fit$p.value,
        n = length(indices)
    )
}

correlation_result <- dig(
    x = dig_iris,
    f = correlation_callback,
    condition = everything(),
    disjoint = disj,
    min_length = 1,
    max_length = 2,
    min_support = 0.1
) |>
    bind_rows() |>
    arrange(desc(abs(correlation)))

head(correlation_result, n = 6)
```

This example follows the same idea as `dig_correlations()`:

- generate conditions,
- evaluate a statistic on the sub-data induced by each condition,
- return one row per successful evaluation.

The difference is that `dig()` leaves the statistic entirely in your hands.
That is useful when you want to fix the variables, apply a custom test, return
additional diagnostics, or combine several criteria in one callback.



# Practical Notes

- Use `var_names(colnames(x))` to build `disjoint` for data prepared with
  `partition()`.
- Use `excluded` together with `dig_tautologies()` and `parse_condition()` when
  you want to prune redundant conditions implied by known axioms.
- For fuzzy searches, request `weights` in the callback instead of `indices`.
- The result of `dig()` is a list. When each callback returns one named list,
  `bind_rows()` is a convenient way to flatten it into a tibble.
- If the callback returns multiple patterns per condition, return a list of
  named lists and flatten the result afterwards.



# Summary

`dig()` is the most flexible search interface in `nuggets`. It lets you:

1. generate conditions from selected predicates,
2. optionally evaluate focus predicates within each condition,
3. receive only the callback inputs you need,
4. define your own pattern logic, statistics, and output format.

Use `dig()` when the built-in search functions are close to what you need, but
not exact. For related material, see:

- `vignette("data-preparation")` for creating crisp and fuzzy predicates,
- `vignette("association-rules")` for a specialized pattern family based on the
  same search principles,
- `vignette("nuggets")` for an overview of the package and its main workflows.
